---
title: "ProblemSet6&7"
author: "719_SwapnilBanerjee"
date: "2026-03-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 6: Model Selection

#### Consider the “Hitters” data. For loading the data you need to install ISLR library. We wish to predict a baseball player’s salary on the basis of various predictors. So first we need to choose an optimal model.

```{r}
# Load libraries
library(ISLR)
library(leaps) # For subset selection
```

#### 1. Clean the data by removing all rows that have missing value in any variable. You will be working with this clean data.

```{r}
# Load data
data(Hitters)

# Remove rows with missing values
hitters <- na.omit(Hitters)

# Check dimensions
dim(hitters)
```

##### Comments

The dataset initially contains some missing values in the Salary variable. These observations are removed using na.omit() to ensure that the analysis is performed only on complete cases.

After cleaning the dataset, the number of observations is reduced but the dataset becomes suitable for regression modelling without missing data issues.

#### 2. Split the data randomly into train-test keeping 80% as training sample and remaining as test sample.

```{r}
set.seed(123)

n <- nrow(hitters)
train_index <- sample(1:n, 0.8*n)

train <- hitters[train_index, ]
test <- hitters[-train_index, ]
```

##### Comments

The dataset is randomly divided into:

80% training data: used to build the regression models. 20% test data: used to evaluate predictive performance.

Setting a seed ensures that the random split is reproducible.

#### 3. Perform the best subset selection on the training sample. While choosing the best model fit up to a 19 variable model. Choose the best model according to adjusted R squared, BIC , AIC and Cp values. In case the best model differs according to the selection criteria, report the best fitted model according to each selection criterion.

```{r}
best_model <- regsubsets(Salary ~ ., data=train, nvmax=19)

best_summary <- summary(best_model)

#Extract Selection Criteria
adjr2 <- best_summary$adjr2
cp <- best_summary$cp
bic <- best_summary$bic

# AIC calculation
rss <- best_summary$rss
n_train <- nrow(train)
p <- 1:19

aic <- n_train*log(rss/n_train) + 2*p

#Best Models According to Criteria
best_adjr2 <- which.max(adjr2)
best_cp <- which.min(cp)
best_bic <- which.min(bic)
best_aic <- which.min(aic)

best_adjr2
best_cp
best_bic
best_aic

#Variables in best models
coef(best_model, best_adjr2)
coef(best_model, best_cp)
coef(best_model, best_bic)
coef(best_model, best_aic)

#Prediction function needed for Test MSE.
predict_regsubsets <- function(object, newdata, id){
  
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id=id)
  xvars <- names(coefi)
  
  mat[,xvars] %*% coefi
}
```

##### Interpretation

Best subset selection evaluates all possible combinations of predictors and selects the best model for each model size. This method guarantees finding the optimal subset of predictors but can be computationally expensive.

Each model selection criterion identifies a potentially different optimal model. Adjusted R² selects the model that explains the largest proportion of variance while penalizing unnecessary predictors. Cp identifies models with low bias and variance. BIC heavily penalizes model complexity, favoring simpler models. AIC balances goodness of fit with model complexity.

If these criteria select different models, it indicates that there is a trade-off between model complexity and predictive accuracy.

#### 4. Find the training MSE and test MSE for each of the best models selected according to different criterion as mentioned in 3 and compare their values.

```{r}
calc_mse <- function(id){

train_pred <- predict_regsubsets(best_model, train, id)
test_pred <- predict_regsubsets(best_model, test, id)

train_mse <- mean((train$Salary - train_pred)^2)
test_mse <- mean((test$Salary - test_pred)^2)

return(c(train_mse, test_mse))
}

mse_adjr2 <- calc_mse(best_adjr2)
mse_cp <- calc_mse(best_cp)
mse_bic <- calc_mse(best_bic)
mse_aic <- calc_mse(best_aic)

mse_adjr2
mse_cp
mse_bic
mse_aic
```

#### 5. Repeat 3-4 using forward stepwise selection method.

```{r}
forward_model <- regsubsets(Salary ~ ., data=train,
                            nvmax=19, method="forward")

forward_sum <- summary(forward_model)

adjr2_f <- which.max(forward_sum$adjr2)
cp_f <- which.min(forward_sum$cp)
bic_f <- which.min(forward_sum$bic)

rss_f <- forward_sum$rss
aic_f <- n_train*log(rss_f/n_train) + 2*(1:19)
aic_f_best <- which.min(aic_f)
```

##### Forward MSE Calculation.

```{r}
calc_mse_forward <- function(id){

train_pred <- predict_regsubsets(forward_model, train, id)
test_pred <- predict_regsubsets(forward_model, test, id)

c(mean((train$Salary-train_pred)^2),
  mean((test$Salary-test_pred)^2))
}

calc_mse_forward(adjr2_f)
calc_mse_forward(cp_f)
calc_mse_forward(bic_f)
calc_mse_forward(aic_f_best)
```

#### 6. Repeat 3-4 using forward stepwise selection method.

```{r}
backward_model <- regsubsets(Salary ~ ., data=train,
                             nvmax=19, method="backward")

backward_sum <- summary(backward_model)

adjr2_b <- which.max(backward_sum$adjr2)
cp_b <- which.min(backward_sum$cp)
bic_b <- which.min(backward_sum$bic)

rss_b <- backward_sum$rss
aic_b <- n_train*log(rss_b/n_train) + 2*(1:19)
aic_b_best <- which.min(aic_b)
```

##### Backward MSE Calculation.

```{r}
calc_mse_backward <- function(id){

train_pred <- predict_regsubsets(backward_model, train, id)
test_pred <- predict_regsubsets(backward_model, test, id)

c(mean((train$Salary-train_pred)^2),
  mean((test$Salary-test_pred)^2))
}

calc_mse_backward(adjr2_b)
calc_mse_backward(cp_b)
calc_mse_backward(bic_b)
calc_mse_backward(aic_b_best)
```

#### 7. Compare the results of the three methods in terms of choice of best model, train MSE amd test MSE.

*Best Subset*\
Evaluates all possible models\
Most computationally expensive\
Often gives best fit\

*Forward Selection*\
Adds predictors gradually\
Faster but may miss optimal model\

*Backward Selection*\
Starts with full model\
Removes unnecessary predictors\


Best subset selection evaluates all possible predictor combinations and generally produces the most optimal model according to statistical criteria such as Adjusted R², BIC, AIC and Cp. Forward and backward stepwise selection are computationally efficient alternatives that approximate the best subset results. The final comparison of training and test MSE indicates which model provides the best predictive performance.


--------------------------------------------------------------------------------------------------


## Problem Set 7: Shrinkage Methods

#### Comparison of least squares, ridge regression, LASSO and Elastic Net

You would be working with Hitters data set available in the ISLR library. You have to predict the Salary based on all available variables. First make sure that you have deleted all rows with one or more missing values.

```{r}
rm(list=ls())
library(ISLR)
data(Hitters)

# remove missing values
hitters = na.omit(Hitters)
dim(hitters)
```
#### 1. Split the data set into a training set and a test set randomly keeping 100 observations in test set and remaining in the training set.
```{r}
set.seed(123)

n = nrow(hitters)

test_index = sample(1:n, 100)

train = hitters[-test_index, ]
test  = hitters[test_index, ]
```


#### 2. Train the model using least squares and report the train MSE and test MSE.

```{r}
library(glmnet)
#model.matrix() converts categorical variables into dummy variables
#[, -1] removes intercept column
x_train = model.matrix(Salary ~ ., train)[,-1]
y_train = train$Salary

x_test = model.matrix(Salary ~ ., test)[,-1]
y_test = test$Salary

#Ridge Model
lm_model = lm(Salary ~ ., data = train)

train_pred = predict(lm_model, train)
test_pred  = predict(lm_model, test)

train_mse_lm = mean((train$Salary - train_pred)^2)
test_mse_lm  = mean((test$Salary - test_pred)^2)

train_mse_lm
test_mse_lm
```
Train MSE → model fit \

Test MSE → prediction accuracy \

#### 3. Fit ridge regression models on the training set, with grid values of λ. Report the following:
```{r}
lambda_grid = 10^seq(5, -2, length = 100)

ridge_model = glmnet(x_train, y_train,alpha = 0,lambda = lambda_grid)
```

*Explanation:*\
alpha = 0 → Ridge regression\
lambda_grid → penalty values\

(i) the dimension of the coefficient matrix so obtained.
```{r}
dim(coef(ridge_model))
```
*Interpretation:*\
rows → predictors + intercept \
columns → number of lambda values \

(ii) The l2 norm of the regression coefficients corresponding to the 50th and 60th choice of λ. Comment on the output.
```{r}
coef50 = coef(ridge_model, s = lambda_grid[50])
coef60 = coef(ridge_model, s = lambda_grid[60])

norm50 = sqrt(sum(coef50[-1]^2))
norm60 = sqrt(sum(coef60[-1]^2))

norm50
norm60
```
Larger λ → stronger shrinkage → smaller L2 norm \

(iii) Test MSE corresponding to these two choices of λ.\
```{r}
pred50 = predict(ridge_model, s = lambda_grid[50], newx = x_test)
pred60 = predict(ridge_model, s = lambda_grid[60], newx = x_test)

mse50 = mean((y_test - pred50)^2)
mse60 = mean((y_test - pred60)^2)

mse50
mse60
```