---
title: "Assignment_5"
author: "Swapnil_Banerjee_719"
date: "2026-02-26"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **PRACTICE SET 3** 

### 3. Problem to demonstrate the role of qualitative (ordinal) predictors in addition to quantitative predictors in multiple linear regression.

Consider “diamonds” data set in R. It is in the ggplot2 package. Make a list of all the ordinal categorical variables. Identify the response.
```{r}
library(ggplot2)
data("diamonds")
View(diamonds)
```
ANSWER: The ordinal categorical variables in the diamonds dataset are "cut", "color", and "clarity". The response variable is "price".

(a) Run a linear regression of the response on the quality of cut. Write the fitted regression model.
```{r}
model_cut <- lm(price ~ cut, data = diamonds)
```

(b) Test whether the expected price of diamond with premium cut is significantly different from that of the ideal cut.
```{r}
summary(model_cut)
```
(c) What is the expected price of a diamond of ideal cut?
```{r}
ideal_price <- predict(model_cut, newdata = data.frame(cut = "Ideal"))
ideal_price
```
(d) Modify the regression model in (a) by incorporating the predictor “table”. Write the fitted regression model.
```{r}
model_cut_table <- lm(price ~ cut + table, data = diamonds)
model_cut_table
```

(e) Test for the significance of “table” in predicting the price of diamond.
```{r}
summary(model_cut_table)
```
(f)  Find the average estimated price of a diamond with an average table value
and which is of fair cut.
```{r}
average_table <- mean(diamonds$table)
average_table
pred= predict(model_cut_table,newdata = data.frame(cut="Fair",table=average_table));pred
```

## **Problem Set 5: K nearest neighbors regression**

```{r}
set.seed(123)
library(class)
#library(caret)
library(FNN)
```


### Q1. Problem to demonstrate the utility of K nearest neighbour regression over least squares regression.

#### Consider a setting with n = 1000 observations. Generate

##### (i) x1i from N(0,22) and x2i from Poisson(λ = 1.5).

##### (ii) ϵi from N(0,1).

##### (iii) yi = −2 +1.4x1i −2.6x2i +ϵi.
```{r}
set.seed(123)

n <- 1000

x1 <- rnorm(n, mean = 0, sd = 2)
x2 <- rpois(n, lambda = 1.5)
epsilon <- rnorm(n, mean = 0, sd = 1)

y <- -2 + 1.4*x1 - 2.6*x2 + epsilon

data <- data.frame(x1, x2, y)
```


##### Split the data into train and test sets. Keep the first 800 observations as training data and the remaining as test data. 
```{r}
train_data <- data[1:800, ]
test_data  <- data[801:1000, ]
```


##### Work out the following:
###### 1. Fit a multiple linear regression equation of y on x1 and x2. Calculate test MSE.
```{r}
lm_model <- lm(y ~ x1 + x2, data = train_data)

summary(lm_model)
```
```{r}
pred_lm <- predict(lm_model, newdata = test_data)

mse_lm <- mean((test_data$y - pred_lm)^2)

mse_lm
```


##### 2. Fit a KNN model with k = 1,2,5,9,15. Calculate test MSE for each choice of k.
```{r}
library(FNN)

train_x <- train_data[, c("x1", "x2")]
test_x  <- test_data[, c("x1", "x2")]
train_y <- train_data$y
```

```{r k=1}
pred_k1 <- knn.reg(train = train_x, test = test_x, y = train_y, k = 1)$pred

mse_k1 <- mean((test_data$y - pred_k1)^2)
mse_k1
```

```{r k=2}
pred_k2 <- knn.reg(train = train_x, test = test_x, y = train_y, k = 2)$pred

mse_k2 <- mean((test_data$y - pred_k2)^2)
mse_k2
```

```{r k=5}
pred_k5 <- knn.reg(train = train_x, test = test_x, y = train_y, k = 5)$pred

mse_k5 <- mean((test_data$y - pred_k5)^2)
mse_k5
```

```{r k=9}
pred_k9 <- knn.reg(train = train_x, test = test_x, y = train_y, k = 9)$pred

mse_k9 <- mean((test_data$y - pred_k9)^2)
mse_k9
```

```{r k=15}
pred_k15 <- knn.reg(train = train_x, test = test_x, y = train_y, k = 15)$pred

mse_k15 <- mean((test_data$y - pred_k15)^2)
mse_k15
```

##### Suppose the data in Step (iii) is generated as :
##### yi = 1/(−2+1.4x1i −2.6x2i +2.9x21i) + 3.1sin(x2i) − 1.5x1ix22i + ϵi.

##### Work out the problems in (1) and (2). Compare and comment on the results.

```{r}
epsilon2 <- rnorm(n, mean = 0, sd = 1)

y_nl <- 1/(-2 + 1.4*x1 - 2.6*x2 + 2.9*(x1^2)) + 3.1*sin(x2) - 1.5*x1*(x2^2) + epsilon2

data_nl <- data.frame(x1, x2, y = y_nl)
```

```{r rain-test split}
train_nl <- data_nl[1:800, ]
test_nl  <- data_nl[801:1000, ]
```

```{r}
lm_model_nl <- lm(y ~ x1 + x2, data = train_nl)

summary(lm_model_nl)
```

```{r}
pred_lm_nl <- predict(lm_model_nl, newdata = test_nl)

mse_lm_nl <- mean((test_nl$y - pred_lm_nl)^2)

mse_lm_nl
```

```{r}
train_x_nl <- train_nl[, c("x1", "x2")]
test_x_nl  <- test_nl[, c("x1", "x2")]
train_y_nl <- train_nl$y
```

```{r}
pred_k1_nl <- knn.reg(train = train_x_nl, test = test_x_nl, y = train_y_nl, k = 1)$pred

mse_k1_nl <- mean((test_nl$y - pred_k1_nl)^2)
mse_k1_nl
```

```{r}
pred_k2_nl <- knn.reg(train = train_x_nl, test = test_x_nl,
                      y = train_y_nl, k = 2)$pred

mse_k2_nl <- mean((test_nl$y - pred_k2_nl)^2)
mse_k2_nl
```
```{r}
pred_k5_nl <- knn.reg(train = train_x_nl, test = test_x_nl,
                      y = train_y_nl, k = 5)$pred

mse_k5_nl <- mean((test_nl$y - pred_k5_nl)^2)
mse_k5_nl
```

```{r}
pred_k9_nl <- knn.reg(train = train_x_nl, test = test_x_nl,
                      y = train_y_nl, k = 9)$pred

mse_k9_nl <- mean((test_nl$y - pred_k9_nl)^2)
mse_k9_nl
```

```{r}
pred_k15_nl <- knn.reg(train = train_x_nl, test = test_x_nl,
                       y = train_y_nl, k = 15)$pred

mse_k15_nl <- mean((test_nl$y - pred_k15_nl)^2)
mse_k15_nl
```
#### Interpretation:

##### Linear True Model:
1.Linear regression gives lowest Test MSE.
2.KNN slightly worse.
3.Because the true relationship is linear.


##### Nonlinear True Model

1.Linear regression performs poorly (misspecified).
2.KNN performs better for moderate k (usually 5 or 9).
3.Small k → high variance.
4.Large k → high bias.


